### 2021-12-09 documents/sections/verbs.tex
go down trimming

\subsubsection{\rc{ɨpɨtə} \qu{to go down}}
\label{sec:godown}
Reflexes of this verb were not affected by the extensions of \rc{k-} in \PPek \pcref{sec:pekodian} and \obj{k-} in \akuriyo \pcref{sec:akuriyo}.
The resistance against the former extension was subsequently overcome in \bakairi; its fate in \ikpeng is unknown.
When \akuriyo extended \obj{k-}, the verb already had an irregularly inflected first person form with \obj{p-}, inherited from \PTir.
At first sight, it may seem that it was also affected by the two independent extensions of \obj{j-} in \carijo \exref{gowhere.car-32} and \yukpa \exref{gowhere.yuk-14}.

\pex<gowhere>
\a<car-32> \carijo \perscommpar{David Felipe Guerrero}\\
\begingl
\gla irə wat͡ʃinakano tae j-ehɨtə-e//
\glb \gl{inan}.\gl{ana} body.of.water along.bounded \gl{1}-go.down-\gl{npst}//
\glft \qu{…I go down through that guachinacán.}//
\endgl
\a<yuk-14> \yukpa \parencite[][]{meira2003primeras}\\
\begingl
\glpreamble aw yéwtu//
\gla aw j-ewuhtu//
\glb \gl{1}\gl{pro} \gl{1}-go.down//
\glft \qu{I went down.}//
\endgl
\xe
%
However, a broader comparative perspective reveals a much more complicated story.
\cref{tab:godown} shows all attested reflexes of \rc{ɨpɨtə}, including verb class membership where applicable; parenthesized forms indicate uncertainty about cognacy status.%
\footnote{The reconstruction of \PPek \rc{ɨptə} in \cref{tab:godown} treats the suffixed elements found in the daughter languages as non-cognate.
\textcite{meira2005southern} make no mention of a regular correspondence between \bakairi \obj{gɨ} and \ikpeng \obj{ŋ}, and at least the addition of a final \obj{ŋ} in \PXin is attested elsewhere:
		\begin{inlinelist}
			\item \PC \rc{əne} \qu{to see}, \arara and \ikpeng \obj{eneŋ} %\parencites[8]{gildea2007greenberg}[56]{alves2017arara}[25]{ikpengpacheco2001}
			\item \PC \rc{əta} \qu{to hear}, \arara \obj{taŋ}, \ikpeng \obj{iraŋ} %\parencites[8]{gildea2007greenberg}[144]{alves2017arara}[270]{ikpengpacheco2001}
			\item \PC \rc{ənə} \qu{to eat meat}, \arara \obj{oŋoŋ} \qu{to bite} %\parencites[8]{gildea2007greenberg}[57]{alves2017arara}
		\end{inlinelist} \parencites[8]{gildea2007greenberg}[56, 144, 57]{alves2017arara}[25, 270]{ikpengpacheco2001}.}
It turns out that while the form \rc{ɨpɨtə} can be reconstructed to \PC, different (proto-)languages do not agree about the class of this verb.
Its reflexes in languages that preserve the split-\gl{s} system are distributed fairly evenly between \gl{s_a_} and \gl{s_p_}.

@@

In one language, \wayana, the verb shows traits of both classes, necessitating an analysis as a \dbqu{mixed} verb in a synchronic description of \wayana.
It takes the first and second person \gl{s_p_} markers \obj{j-} and \obj{əw-} \parencite[200]{wayanatavares2005}, but the \gl{1+2}\gl{s_a_} marker \obj{kut-} \parencite[206]{wayanatavares2005}.
It also shows the \gl{s_a_} class marker \obj{w-} in nominalizations \exref{waygodown.way-73}, but behaves like an \gl{s_p_} verb in taking a second person prefix in imperatives \exref{waygodown.way-71}.

\pex<waygodown>\wayana \parencite[][200]{wayanatavares2005}
\a<way-73>
\begingl
\glpreamble ïwïptëë//
\gla ɨ-w-ɨptə-rɨ//
\glb \gl{1}-\gl{s_a_}-go.down-\gl{nmlz}//
\glft \qu{my going down}//
\endgl
\a<way-71>
\begingl
\gla əw-ɨptə-k//
\glb \gl{2}-go.down-\gl{imp}//
\glft \qu{Go down!}//
\endgl
\xe
%
Its causativized form is \obj{ɨptə-ka} \parencite[255]{wayanatavares2005}; the restriction of \rc{-ka} to \gl{s_p_} verbs in \PC \parencite{gildea2019overview} points to \gl{s_p_} membership.
These patterns are suggestive of a scenario in which the verb was a regular member of the \gl{s_p_} class in pre-\wayana, but partially switched to the \gl{s_a_} class, reflected only in its \gl{1+2} prefix and the \gl{s_a_} class marker.
This in turn implies that reflexes of this verb with \gl{s_a_} membership in other languages fully switched from \gl{s_p_} in their inflectional patterns.

\wayana-external comparative evidence supports this hypothesis:
The \arara causativized form is \obj{eniptoŋ} \parencite[66]{alves2017arara}, and \kalina has a cognate form \obj{enɨʔto} \parencite[263]{courtz2008carib}; \obj{onɨʔto} \qu{to go down} in \cref{tab:godown} is a detransitivized form thereof, lit.\ \qu{to get oneself down}.
Both causativized forms contain a reflex of the transitivizer \rc{en-}, which was usually found with \gl{s_p_} verbs \parencite{gildea2019overview}.
Besides the irregular first person \obj{p-}, \trio \obj{ɨhtə} shows other irregularities, in particular in its causativized forms that also feature a reflex of \rc{en-} \parencite[263]{triomeira1999}.
Thus, it appears that this verb was originally \gl{s_p_}, but then switched its class in four and a half languages of the family, for so far unknown reasons.

These circumstances make it impossible to answer the question of whether \qu{to go down} was affected by the extensions in \PWai, \PTir, \carijo, and \yukpa.
For \PTir, one cannot establish a relative chronology of the verb class switch, the introduction of the idiosyncratic marker \rc{p-}, and the extension of \rc{t-}.
For \PWai, not only the first person form of the verb is unknown, but even its class membership.
Similarly, for \carijo and \yukpa one cannot say whether the verb potentially switched class before the breakdown of the entire split-\gl{s} system.
While there is no language-internal evidence for such a switch, it is clear that \qu{to go down} does have an inclination to switch classes; in the case of \carijo, that could have already happened at the \PTar stage.
In all four cases, it is possible that the verb had \gl{s_a_} status at the time of the extension, resisting it and preserving the old prefix, but it is also possible that it was not even a potential target due to its \gl{s_p_} status at the time.

On the other hand, the class switch happened before the split-up of \trio and \akuriyo, and therefore this verb resisted the extension of \akuriyo \obj{k-} as an \gl{s_a_} verb.
Likewise, it is very likely that the class switch took place before the extension of \PPek \rc{k-}.
Otherwise, the newly-turned-\gl{s_a_} verb would have taken on conservative and lexically heavily restricted \rc{w-}, either in \PPek, \PXin, or \arara.

>>>>

\subsubsection{\rc{ɨpɨtə} \qu{to go down}}
\label{sec:godown}
Reflexes of this verb were not affected by the extensions of \rc{k-} in \PPek \pcref{sec:pekodian} and \obj{k-} in \akuriyo \pcref{sec:akuriyo}.
The resistance against the former extension was subsequently overcome in \bakairi; its fate in \ikpeng is unknown.
When \akuriyo extended \obj{k-}, the verb already had a first person form irregularly inflected with \obj{p-}, inherited from \PTir.
One might think that it was also affected by the extensions of \obj{j-} in \carijo \exref{gowhere.car-32} and \yukpa \exref{gowhere.yuk-14}.

\pex<gowhere>
\a<car-32> \carijo \perscommpar{David Felipe Guerrero}\\
\begingl
\gla irə wat͡ʃinakano tae j-ehɨtə-e//
\glb \gl{inan}.\gl{ana} body.of.water along.bounded \gl{1}-go.down-\gl{npst}//
\glft \qu{…I go down through that guachinacán.}//
\endgl
\a<yuk-14> \yukpa \parencite[][]{meira2003primeras}\\
\begingl
\glpreamble aw yéwtu//
\gla aw j-ewuhtu//
\glb \gl{1}\gl{pro} \gl{1}-go.down//
\glft \qu{I went down.}//
\endgl
\xe
%
However, a broader comparative perspective reveals a much more complicated story \pcref{tab:godown}.%
\footnote{In \cref{tab:godown}, parenthesized forms indicate uncertainty about cognacy status. The reconstruction of \PPek \rc{ɨptə} treats the suffixed elements found in the daughter languages as non-cognate.
\textcite{meira2005southern} identify no correspondence between \bakairi \obj{gɨ} and \ikpeng \obj{ŋ}, and at least the addition of a final \obj{ŋ} in \PXin is attested elsewhere:
		\begin{inlinelist}
			\item \PC \rc{əne} \qu{to see}, \arara and \ikpeng \obj{eneŋ} %\parencites[8]{gildea2007greenberg}[56]{alves2017arara}[25]{ikpengpacheco2001}
			\item \PC \rc{əta} \qu{to hear}, \arara \obj{taŋ}, \ikpeng \obj{iraŋ} %\parencites[8]{gildea2007greenberg}[144]{alves2017arara}[270]{ikpengpacheco2001}
			\item \PC \rc{ənə} \qu{to eat meat}, \arara \obj{oŋoŋ} \qu{to bite} %\parencites[8]{gildea2007greenberg}[57]{alves2017arara}
		\end{inlinelist} \parencites[8]{gildea2007greenberg}[56, 144, 57]{alves2017arara}[25, 270]{ikpengpacheco2001}.}
It turns out that while a verb\rc{ɨpɨtə} can be reconstructed to \PC, different (proto-)languages do not agree about its class.
Its reflexes in languages that preserve the split-\gl{s} system are distributed fairly evenly between \gl{s_a_} and \gl{s_p_}.

@@

The verb shows traits of both classes in \wayana, necessitating an analysis as a \dbqu{mixed} verb in a synchronic description of that language.
It takes the first and second person \gl{s_p_} markers \obj{j-} and \obj{əw-} \parencite[200]{wayanatavares2005}, but the \gl{1+2}\gl{s_a_} marker \obj{kut-} \parencite[206]{wayanatavares2005}.
It also shows the \gl{s_a_} class marker \obj{w-} in nominalizations \exref{waygodown.way-73}, but behaves like an \gl{s_p_} verb in taking a second person prefix in imperatives \exref{waygodown.way-71}.

\pex<waygodown>\wayana \parencite[][200]{wayanatavares2005}
\a<way-73>
\begingl
\glpreamble ïwïptëë//
\gla ɨ-w-ɨptə-rɨ//
\glb \gl{1}-\gl{s_a_}-go.down-\gl{nmlz}//
\glft \qu{my going down}//
\endgl
\a<way-71>
\begingl
\gla əw-ɨptə-k//
\glb \gl{2}-go.down-\gl{imp}//
\glft \qu{Go down!}//
\endgl
\xe
%
Its causativized form is \obj{ɨptə-ka} \parencite[255]{wayanatavares2005}; the \PC  causativizer \rc{-ka} was restricted to \gl{s_p_} verbs \parencite{gildea2019overview}.
These patterns point to a scenario where the verb was a regular member of the \gl{s_p_} class in pre-\wayana, but partially switched to the \gl{s_a_} class, taking a \gl{1+2}\gl{s_a_} prefix and the \gl{s_a_} class marker.
This in turn implies that \gl{s_a_} reflexes of this verb in other languages fully switched from \gl{s_p_} in their inflectional patterns.

\wayana-external comparative evidence supports this hypothesis:
The \arara causativized form is \obj{eniptoŋ} \parencite[66]{alves2017arara}, and \kalina has a cognate form \obj{enɨʔto} \parencite[263]{courtz2008carib}; \obj{onɨʔto} \qu{to go down} in \cref{tab:godown} is a detransitivized form thereof, lit.\ \qu{to get oneself down}.
Both causativized forms contain a reflex of the transitivizer \rc{en-}, which was usually found with \gl{s_p_} verbs \parencite{gildea2019overview}.
\trio \obj{ɨhtə} has irregular causativized forms that also feature a reflex of \rc{en-} \parencite[263]{triomeira1999}.
Thus, it appears that this verb was originally \gl{s_p_}, but then switched its class in four \dbqu{and a half} languages of the family, for so far unknown reasons.

These circumstances make it impossible to answer the question of whether \qu{to go down} was affected by some extensions.
For \PTir, one cannot establish a relative chronology of the verb class switch, the introduction of the idiosyncratic marker \rc{p-}, and the extension of \rc{t-}.
For \PWai, both its first person form and its inflectional class are unknown.
For \carijo and \yukpa, one cannot rule out the possibility that the verb switched classes before the breakdown of the split-\gl{s} system.
While no language-internal evidence supports that scenario, \qu{to go down} clearly has an inclination to switch classes; in the case of \carijo, that could have already happened at the \PTar stage.
In all four cases, the verb could have had \gl{s_a_} status at the time of the extension, resisting it and preserving the old prefix, but it could also have had \gl{s_p_} status and thus not even have been a potential target.

On the other hand, the class switch happened before the split of \PTir, and therefore this verb resisted the extension of \akuriyo \obj{k-} as an \gl{s_a_} verb.
Likewise, it is very likely that the class switch took place before the extension of \PPek \rc{k-}.
Otherwise, the newly-turned-\gl{s_a_} verb would have taken on conservative and lexically heavily restricted \rc{w-}, either in \PPek, \PXin, or \arara.
